---
title: "Danish Kings Week 10"
author: "Mathias Dyhr Pedersen"
date: "2025-03-05"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator:komma

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs

library(tidyverse)

kings1 <- read.csv("data/danish_kings_rigtig.csv")

kings2 <- read_csv("data/danish_kings_rigtig.csv")

kings3 <- read.csv2("data/danish_kings_rigtig.csv")

kings4 <- read_csv2( "data/danish_kings_rigtig.csv")
```
Answer: 
1. Which of these functions is a tidyverse function? Read data with it below into a kings object

The "read_csv" functions are tivyverse functions as they reveal a "tibble"

2. What is the result of running class() on the kings object created with a tidyverse function.

When I ran the functions with "." it reveals a data frame, whereas the functions with "_" reveals a tibble. And further, the dataset recognizes the data as characters when using the "read.csv" functions, whereas the "read_csv" functions recognizes the data as digits. 

3. How many columns does the object have when created with these different functions?

The objects which I created using "read.csv" and "read_csv" contains 5 columns because my data is seperated by commas. However the objects which I created using "read.csv2" and "read_csv2) contains 1 column because my data is not seperated by semicolons.

4. Show the dataset so that we can see how R interprets each column

See following functions below

```{r eval=TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- kings2

class(kings)


glimpse(kings)


head(kings)


```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r eval=TRUE}
kings_no_na <- kings2 %>% 
  filter(!is.na(sidste_regeringsår))

kings_duration <- kings_no_na %>% 
  mutate(duration = sidste_regeringsår - første_regeringsår)


```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r eval=TRUE}
kings_no_na <- kings2 %>% 
  filter(!is.na(sidste_regeringsår))

kings_duration <- kings_no_na %>% 
  mutate(duration = sidste_regeringsår - første_regeringsår)

kings_average_duration <- kings_duration$duration %>% 
  mean( ,kings_duration$duration)
```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r eval=TRUE}
kings_long_regin <- kings_duration %>% 
  filter(duration>kings_average_duration)

count(kings_long_regin)
```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r eval=TRUE}
kings_longest_reign <- kings_duration %>% 
  select(duration, regent) %>% 
  arrange(desc(duration)) %>% 
  slice_max(order_by = duration, n = 3)

kings_longest_reign %>% 
  mutate(days = duration * 365 + floor(duration / 4))
```
## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r eval=TRUE}
kings_graph <- kings_duration %>% 
  mutate(midyear = første_regeringsår + (sidste_regeringsår - første_regeringsår / 2))

ggplot(kings_graph, mapping = aes(x = midyear, y = duration))+
  geom_point(size = 3)+
  geom_line(colour = "purple")+
  geom_smooth(method = lm, se = FALSE)
  

```


And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


